---
title: "Session 1 — Introduction to R and RStudio"
subtitle: "Applied Methods · MA Public Administration"
author: "Maxime Walder"
date: "September 18, 2026"
format:
  revealjs:
    theme: simple
    embed-resources: true
    slide-number: true
    transition: slide
execute:
  echo: true
  warning: false
  message: false
---

## Today’s objectives

By the end of this first session, you should be able to:

-   explain how the course is organized and evaluated
-   identify the main course objectives and expected outputs
-   form a group and select a first group-project topic
-   install R and RStudio, or verify that they work on your computer
-   use RStudio to run simple R commands in a script and in the console
-   create objects, vectors, factors, functions, and a basic data frame
-   understand the path and relative path structure for CLI-like commands
-   import a first `.csv` data file and inspect it

## But first, who are we?

-   First and last name
-   Current cursus
-   Expectation from this class

<!-- ## Agenda for today -->

<!-- | Time | Activity | -->

<!-- |---|---| -->

<!-- | 10:15–10:45 | Course presentation and learning objectives | -->

<!-- | 10:45–11:20 | Syllabus, evaluation, assignments, group project | -->

<!-- | 11:20–12:00 | Group formation and topic choice | -->

<!-- | 12:00–13:00 | Lunch break | -->

<!-- | 13:00–13:40 | Installing R and RStudio; first tour of RStudio | -->

<!-- | 13:40–14:25 | R basics: objects, operations, vectors, data frames, functions | -->

<!-- | 14:25–14:50 | Guided exercises | -->

<!-- | 14:50–15:00 | Wrap-up and preparation for Session 2 | -->

## The course in one sentence

This course teaches you how to use R to transform public and administrative data into reproducible, interpretable, and communicable policy analysis.

## Course philosophy

We will not start from abstract formulas.

We will start from practical questions:

-   Where are the data?
-   What is the unit of analysis?
-   What needs to be cleaned, recoded, summarized, or visualized?
-   Which result answers the policy question?
-   Can someone else reproduce the analysis?

## What you will learn during the semester

-   Data import, cleaning, and transformation
-   Descriptive statistics and data visualization
-   Hypothesis testing and regression models
-   Interaction terms and predicted values
-   Quarto reports
-   Shiny applications
-   Introductory text-as-data methods

## Session planning I

| Session | Date     | Topic                                        |
|--------:|----------|----------------------------------------------|
|       1 | Sept. 18 | Introduction to R and RStudio                |
|       2 | Sept. 25 | Data manipulation and descriptive statistics |
|       3 | Oct. 2   | Data visualization                           |
|      \- | Oct. 9   | *Dies academic*                              |
|       4 | Oct. 16  | Assignment 1 + Quarto reports                |
|       5 | Oct. 23  | Hypothesis testing and regression            |
|       6 | Oct. 30  | Interaction terms                            |

## Session planning II

| Session | Date         | Topic                         |
|--------:|--------------|-------------------------------|
|      \- | Nov. 6       | *Reading week*                |
|       7 | Nov. 13      | Assignment 2 + project coding |
|       8 | Nov. 20      | Text as data                  |
|       9 | Nov. 27      | Shiny                         |
|      10 | Dec. 4       | Assignment 3 + web scraping   |
|   11–12 | Dec. 11 & 18 | Group project presentations   |

## Evaluation: overview

The final grade combines three complementary components.

| Component | Output | Role |
|------------------------|------------------------|------------------------|
| In-class assignments | 3 problem sets | Practice and individual assessment |
| Group project | R script, Quarto report, Shiny app, presentation | Applied policy analysis |
| Oral exam | Individual discussion of your own project | Code, interpretation, and justification |

## Assignment 1

**Date:** October 16

Focus:

-   inspect a dataset
-   manipulate and recode variables
-   compute descriptive statistics
-   produce basic visualizations
-   submit a reproducible R script

## Assignment 2

**Date:** November 13

Focus:

-   estimate regression models
-   include interaction terms
-   visualize and interpret model results
-   connect empirical results to a substantive policy question

## Assignment 3

**Date:** December 4

Focus:

-   build a small Shiny application
-   display descriptive or model-based results interactively
-   make the app understandable for non-technical users

## Group project: purpose

The group project is a reproducible, data-driven policy report using public or administrative data.

The aim is not only to “run code,” but to produce an analysis that could be useful for public administration.

## Group project: final outputs

Each group submits:

-   a reproducible R script and folder structure
-   a Quarto report: `.qmd` and rendered `.html`
-   a small Shiny application
-   a short in-class work-in-progress presentation

**Final deadline:** January 13, 2027, 20:00.

## Group project: possible data sources

Examples of useful sources:

-   OCSTAT data for Geneva
-   Swiss Federal Statistical Office data
-   opendata.swiss datasets
-   municipal, cantonal, or federal administrative data
-   public policy documents or parliamentary texts
-   any other data source, be original!

## Choosing a topic today

A good topic should be:

-   narrow enough to be feasible in one semester
-   connected to a public-administration question
-   based on accessible data
-   answerable with descriptive statistics, visualizations, and possibly simple models
-   suitable for a small Shiny application

## Topics for group project

Group project can tackle any topic. 

However, topic ust be possible to study using quantitative methods. 

Personal suggestion is to limit to existing administrative data.

Let's see what data cube are easily availble with the [OCSTAT](https://statistique.ge.ch/domaines/tableaux.asp)

## Group formation activity

In groups of 3–4 students:

1.  Choose one provisional topic.
2.  Identify the possible unit of analysis: commune, canton, year, individual, organization, document.
3.  Identify one possible outcome variable.
4.  Identify two possible explanatory or descriptive variables.
5.  Write one sentence beginning with:\
    **“Our project studies whether/how...”**

## First feedback group projects

1.  Present the general idea
2.  Explain what you think should be the outcome
3.  Explain what are the explanatory factors

## What is R?

R is a programming language and software environment for statistical computing, data analysis, and visualization.

In this course, R is the engine that performs the analysis.

## Install R

Recommended steps:

1.  Go to CRAN: <https://cran.r-project.org/>
2.  Choose your operating system: Windows, macOS, or Linux
3.  Download and install the most recent stable version of R
4.  Open R once to verify that it starts

Note: install **R before RStudio**.

## What is RStudio?

RStudio is an integrated development environment, or IDE.

It makes R easier to use by bringing together:

-   the console
-   script editor
-   environment viewer
-   files, plots, packages, and help
-   project management tools

## Install RStudio

Recommended steps:

1.  Go to Posit’s RStudio download page: <https://posit.co/download/rstudio-desktop/>
2.  Download RStudio Desktop for your operating system
3.  Install and open it
4.  Check that RStudio detects your R installation

## RStudio interface

The four main panes are:

| Pane | What it is for |
|------------------------------------|------------------------------------|
| Source | Write and save scripts |
| Console | Run commands interactively |
| Environment | See objects currently stored in memory |
| Files / Plots / Packages / Help | Navigate files, view plots, load packages, read help |

## Console versus script

The console is useful for testing.

The script is useful for reproducibility.

In this course, most of your work should be written in scripts or Quarto documents, not only typed in the console.

## Classe's workflow

-   All documents are available on Moodle.

-   But also on github [maxwaldo/AppliedMethods](https://github.com/maxwaldo/AppliedMethods)

-   Before each session, you can download the repository and work directly on it.

-   Let's try this!

## Exercice 0

## First basic commands

```{r}
2 + 2
10 - 3
5 * 4
20 / 5
sqrt(16)
```

## Exercice 1

## Assignment operators

In R, we usually store values with `<-`.

```{r}
age <- 24
name <- "Geneva"
ratio <- 10 / 4
```

Read `<-` as “gets”.

`age <- 24` means: the object `age` gets the value `24`.

## Exercice 2

## Objects in the environment

After creating objects, they appear in the Environment pane.

```{r}
number_students <- 28
course_name <- "Applied Methods"
is_required <- TRUE
```

Objects can be reused later.

```{r}
number_students + 2
```

## Object classes

R objects have classes.

```{r}
one_number <- 10
one_character <- "public administration"
one_logical <- TRUE
one_factor <- factor("Geneva")

class(one_number)
class(one_character)
class(one_logical)
class(one_factor)
```

## Exercice 3

## Errors and warnings

Not everything is possible to do in R.

```{r, eval=FALSE}
my_pet <- "Cat"
age_pet <- 4

my_pet+age_pet
```

Why this cannot be done?

Not all command generate an error, some may issue warnings.

## Reading errors

Errors are not failures. They are information.

When you get an error:

1.  read the first line carefully
2.  identify the object or function mentioned
3.  check spelling and capitalization
4.  check whether the package is loaded
5.  run the code line by line


## Exercice 4

## Vectors

A vector stores several values of the same basic type.

In R, you can specify vectors inside of the `c()`

Each element of the vector is separated by a coma.

```{r}
communes <- c("Carouge", "Cologny", "Lancy", "Meyrin")
population <- c(22600, 5700, 34400, 27000)

# You can then perform operations on these vectors
average_population <- mean(population)
average_population
```

## Vectors

Vectors are unidimensional (VS data frame or matrix).

You can access content of vectors by specifying the index of the observation

```{r}
communes[2] # Get the second element of the vector
```

## Exercice 5

## Data frames

A data frame is a rectangular dataset:

-   rows are observations
-   columns are variables

```{r}
municipal_data <- data.frame(
  commune = communes,
  population = population
)

municipal_data
```

## Data frames

Data frames have two dimensions, rows and columns.

You can access the content of specific cells by specifying the index for the rows, and the columns

`data[row,column]`

```{r}

municipal_data[1,] ## First row all columns
municipal_data[,1] ## First column all rows
municipal_data[1,1] ## First row and first column


```

## Data frames

In data frames, the columns are the variables, the lines are the observations.

You can also access the content of specific variable using the atomic character `$`

```{r}
municipal_data$population
```

## Exercice 6

## Functions

A function takes inputs, performs operations, and returns an output.

To create a function, you have a built in `function` option.

```{r}
divide_by_4_add_1 <- function(x) {
  x / 4 + 1
}

divide_by_4_add_1(20)
```

## Functions

You can also have multiple arguments in a function

```{r}
divide_by_4_add_n <- function(x, n) {
  x / 4 + n
}

divide_by_4_add_n(20, 4)
```

## Functions

Functions can also have default values

```{r}
divide_by_4_add_n <- function(x = 10, n = 4) {
  x / 4 + n
}

divide_by_4_add_n(20, 2) # Same as before
divide_by_4_add_n() # Function with default value
```

## Functions

Finally functions can also be more or less robust and well documented

If I run: `divide_by_4_add_n("Cat", "Dog")` `Error in x/4 : non-numeric argument to binary operator`

```{r}

divide_by_4_add_n <- function(x, n) {
  
  if (!is.numeric(x) | !is.numeric(n)) {
    stop(
      paste0(
        "x and n must be numeric, but an object of class non-numeric was provided. \n x is '",
        class(x)[1], 
        "'\n and n is '",
        class(n)[1], "'"
      )
    )
  }
  
  x / 4 + n
}

```

## Exercice 7

## Packages and libraries

Packages extend what R can do.

Install a package once:

```{r, eval=FALSE}
install.packages("readr")
```

Load it in each new R session:

```{r, eval=FALSE}
library(readr)
```

Careful: For some reason, you need to put "" when installing package, but not when loading libraries.

## Packages and libraries

There are lots of packages and libraries functionning in R.

You can also developp your own package and make it available for everyone.

Some packages are available not with the official distribution channel. You can use function from the `devtools` or the `remotes` packages to install these softwares. 

## Exercice 8

## Understanding path and relative path

With R, and other programming languages, it is useful to understand how to navigate your computer. 

Basicall, your computer is organized in folder. You can specify where files and folders are with _paths_.

For instance, the command `getwd()` gives you the current working directory. 

You can change this directory with `setwd()` or you can precise path with caracter variable.



## Importing a first data frame

```{r, eval=FALSE}
# Import with the full path
first_data <- read.csv("~/Desktop/Teaching_Grading/Applied methods/session_01_applied_methods/data/session_01_first_dataset.csv") 

# Or set working directory and import with relative path
setwd("~/Desktop/Teaching_Grading/Applied methods/session_01_applied_methods/")

first_data <- read.csv("data/session_01_first_dataset.csv") 


View(first_data)
head(first_data)
str(first_data)
summary(first_data)
```

## Exercice 9



## Wrap-up

Today you learned how to:

-   navigate the course organization and evaluation
-   form groups and choose a provisional project topic
-   distinguish R from RStudio
-   use the RStudio interface
-   write reproducible R code in a script
-   create objects, vectors, factors, functions, and data frames
-   import and inspect a first dataset

## Questions ?
